---
title: "SDoH 2025 Update"
author: "Carina Korcel"
date: "2025-02-03"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load libraries}
library(dplyr)
library(censusapi)
library(readxl)
library(tidyr)
library(ggplot2)
library(httr)
library(tidyverse)
```

# Data Sources

  1. American Community Survey (ACS) - Survey data from 3.5 million addresses on 40+ topics conducted by the Census Bureau. Topics include age, children, veterans, commuting, education, income, and employment. The data used here are 5-year estimates. New data is available every year. Data reflects 5-year period not a point in time. Data is collected through internet, mail, and in-person visits.
  
  2. CDC Local Data for Better Health (CDC) - Estimates for 36 health measures designed to be used to identify emerging health problems and to help develop and carry out effective, targeted public health prevention activities.
  
  3. CDC Social Vulnerability Index (SVI) - Social vulnerability is the degree to which a community exhibits certain social conditions that may affect that community's ability to prevent human suffering and financial loss in the event of a disaster. Index indicating the relative vulnerability of every U.S. ZCTA. Ranks on 5 social factors and groups them into 4 related themes.
  
  4. National Center for Statistics Urban-Rural Classification Scheme for Counties (NCHS) - Used to study and monitor health disparities across the urban-rural continuum. 6 levels ( 1 - large central metro, 2 - large fringe metro, 3 - medium metro, 4 - small metro, 5 - counties with a city of 10,000 or more population, 6 - counties without a city of 10,000 or more population). Updated in 2023. Minimal differences in county assignments, largely due to population changes over time or to counties being added or deleted.
  
  5. County Health Rankings & Roadmaps (CHR) - Annual study that ranks the health of nearly every county in the nation across a variety of health outcomes/behaviors.
  
  6. US Census Bureau Decennial Census - Used for the total population estimate.

# Data Pulling
  1. ACS - via api call to Census Bureau
  Vintage: 2023
  
  2. CDC - via excel file downloaded from CDC website for ZCTA and county (https://data.cdc.gov/browse?category=500+Cities+%26+Places&q=2024&sortBy=relevance&tags=places&pageSize=20)
  ZCTA (8/23/24)
  County (12/23/24)
  
  3. SVI - via excel file downloaded from CDC website for ZCTA and county (https://www.atsdr.cdc.gov/place-health/php/svi/svi-data-documentation-download.html)
  ZCTA - 2022 (version 12/11/24)
  County - 2022 (version 8/23/24)
  
  4. Zip code to ZCTA crosswalk - via excel file download from UDS Mapper (http://web.archive.org/web/20231218141557/https://udsmapper.org/zip-code-to-zcta-crosswalk/)
  From Census 2020 geographic boundaries for ZCTAs and the ACS 2017-2022 estimates
  
  5. NCHS - via excel file download from NCHS website
  (https://www.cdc.gov/nchs/data-analysis-tools/urban-rural.html)
  Version: 2023
  
  6. CHR - via excel file download from CHR website
  (https://www.countyhealthrankings.org/health-data/methodology-and-sources/data-documentation)
  Version: 2025
  
  7. Total population - via excel file download from US Census Bureau website
  (https://data.census.gov/table/DECENNIALSF12010.P1?g=010XX00US$0500000)
  Version: 2020
  
##########################################################################################################
########################################## Zip Code Level Table ##########################################
##########################################################################################################

###############################################################
############### American Community Survey (ACS) ###############
###############################################################

# Import and clean ACS data
## change vintage year twice
## run 1 table at a time to see if column heads have changed
##    B02001 - capitalization changed
##    B19013 - changed to 2023 inflation-adjusted dollars
```{r pull ACS data from Census Bureau website and clean}
# establish Census API connection
Sys.setenv(CENSUS_KEY= "KEY") # DO NOT SHARE THIS
readRenviron("~/.Renviron")
Sys.getenv("CENSUS_KEY")

# build function to execute census api call
censusapicall <- function(x){
  
  # get field names
  variables <- listCensusMetadata(
    name="acs/acs5",
    vintage="2023", # TO UPDATE: CHANGE VINTAGE
    type="variables",
    group=x) %>%
    filter(grepl("EA",name)==FALSE,
           grepl("MA",name)==FALSE,
           grepl("M",name)==FALSE)
    
  # select field names and descriptions
  vars <- variables$name
  cols <- variables$label
    
  # pull selected fields
  data <- censusapi::getCensus(
    name="acs/acs5",
    vintage="2023", # TO UPDATE: CHANGE VINTAGE
    
    vars=c("NAME", vars),
    region="zip code tabulation area:*") 
    
  # adding column names to data
  colnames(data) <- c("ZCTA", "Name", cols)
  
  # add table to environment
  assign(x, data, envir = .GlobalEnv)
  
}

# create a data dictionary of select tables
dictionary <- as.data.frame(rbind(
  c("B01001", "population by sex and age"),
  c("B02001", "race"),
  c("B03001", "hispanic/latino"),
  c("B05001", "citizenship"),
  c("B21001", "veteran status"),
  c("B11001", "household type"),
  c("B25104", "monthly housing costs"),
  c("B23001", "employment status"),
  c("B17001", "poverty status"),
  c("B17024", "ratio of income to poverty level"),
  c("B19013", "median household income"),
  c("B19083", "Gini index of income inequality"),
  c("B22003", "food stamps/SNAP"),
  c("B15001", "educational attainment for 18+"),
  c("B08006", "means of transportation to work"),
  c("B27001", "health insurance coverage status"),
  c("B18101", "disability status"),
  c("B16005", "limited English"),
  c("B28011", "no internet")
  ) )
colnames(dictionary) <- c("table", "topic")

# use censusapicall function to pull selected tables
api_calls <- sapply(dictionary$table, censusapicall)

#  1. aggregate the detailed cross-section data
#  2. convert totals into proportions
#  3. combine cleaned tables into 1 ACS table

# population by sex and age
B01001_clean <- B01001 %>%
  mutate(total_pop = `Estimate!!Total:`,
         prop_male = `Estimate!!Total:!!Male:`/`Estimate!!Total:`,
         prop_female = `Estimate!!Total:!!Female:`/`Estimate!!Total:`,
         total_age_under_18 = (  rowSums(B01001[,colnames(B01001)[grepl("under 5 years", colnames(B01001), ignore.case = TRUE)]]) +
                                   rowSums(B01001[,colnames(B01001)[grepl("5 to 9 years", colnames(B01001), ignore.case = TRUE)]]) +
                                   rowSums(B01001[,colnames(B01001)[grepl("10 to 14 years", colnames(B01001), ignore.case = TRUE)]]) +
                                   rowSums(B01001[,colnames(B01001)[grepl("15 to 17 years", colnames(B01001), ignore.case = TRUE)]]) ) ,
         total_age_18_to_64 = (  rowSums(B01001[,colnames(B01001)[grepl("18 and 19 years", colnames(B01001), ignore.case = TRUE)]]) +
                                   rowSums(B01001[,colnames(B01001)[grepl("20 years", colnames(B01001), ignore.case = TRUE)]]) +
                                   rowSums(B01001[,colnames(B01001)[grepl("21 years", colnames(B01001), ignore.case = TRUE)]]) +
                                   rowSums(B01001[,colnames(B01001)[grepl("22 to 24 years", colnames(B01001), ignore.case = TRUE)]]) +
                                   rowSums(B01001[,colnames(B01001)[grepl("25 to 29 years", colnames(B01001), ignore.case = TRUE)]]) +
                                   rowSums(B01001[,colnames(B01001)[grepl("30 to 34 years", colnames(B01001), ignore.case = TRUE)]]) +
                                   rowSums(B01001[,colnames(B01001)[grepl("35 to 39 years", colnames(B01001), ignore.case = TRUE)]]) +
                                   rowSums(B01001[,colnames(B01001)[grepl("40 to 44 years", colnames(B01001), ignore.case = TRUE)]]) +
                                   rowSums(B01001[,colnames(B01001)[grepl("45 to 49 years", colnames(B01001), ignore.case = TRUE)]]) +
                                   rowSums(B01001[,colnames(B01001)[grepl("50 to 54 years", colnames(B01001), ignore.case = TRUE)]]) +
                                   rowSums(B01001[,colnames(B01001)[grepl("55 to 59 years", colnames(B01001), ignore.case = TRUE)]]) +
                                   rowSums(B01001[,colnames(B01001)[grepl("60 and 61 years", colnames(B01001), ignore.case = TRUE)]]) +
                                   rowSums(B01001[,colnames(B01001)[grepl("62 to 64 years", colnames(B01001), ignore.case = TRUE)]]) ) ,
         total_age_over_64 = (  rowSums(B01001[,colnames(B01001)[grepl("65 and 66 years", colnames(B01001), ignore.case = TRUE)]]) +
                                   rowSums(B01001[,colnames(B01001)[grepl("67 to 69 years", colnames(B01001), ignore.case = TRUE)]]) +
                                   rowSums(B01001[,colnames(B01001)[grepl("70 to 74 years", colnames(B01001), ignore.case = TRUE)]]) +
                                   rowSums(B01001[,colnames(B01001)[grepl("75 to 79 years", colnames(B01001), ignore.case = TRUE)]]) +
                                   rowSums(B01001[,colnames(B01001)[grepl("80 to 84 years", colnames(B01001), ignore.case = TRUE)]]) +
                                   rowSums(B01001[,colnames(B01001)[grepl("85 years and over", colnames(B01001), ignore.case = TRUE)]]) ),
         prop_age_under_18 = total_age_under_18/`Estimate!!Total:`,
         prop_age_18_to_64 = total_age_18_to_64/`Estimate!!Total:`,
         prop_age_over_64 = total_age_over_64/`Estimate!!Total:` ) %>%
  select(ZCTA,
         prop_female,
         prop_age_under_18,
         prop_age_18_to_64,
         prop_age_over_64)

# race
B02001_clean <- B02001 %>%
  mutate(total_race = `Estimate!!Total:`,
         prop_race_two_or_more = `Estimate!!Total:!!Two or More Races:`/total_race,
         prop_race_alone_other = `Estimate!!Total:!!Some Other Race alone`/total_race,
         prop_race_alone_white = `Estimate!!Total:!!White alone`/total_race,
         prop_race_alone_HI_PI = `Estimate!!Total:!!Native Hawaiian and Other Pacific Islander alone`/total_race,
         prop_race_alone_asian = `Estimate!!Total:!!Asian alone`/total_race,
         prop_race_alone_AI_AK = `Estimate!!Total:!!American Indian and Alaska Native alone`/total_race,
         prop_race_alone_black = `Estimate!!Total:!!Black or African American alone`/total_race,
         prop_minority = (prop_race_two_or_more +
                          prop_race_alone_other +
                          prop_race_alone_HI_PI +
                          prop_race_alone_asian +
                          prop_race_alone_AI_AK +
                          prop_race_alone_black) ) %>%
  select(ZCTA,
         prop_race_two_or_more,
         prop_race_alone_other,
         prop_race_alone_white,
         prop_race_alone_HI_PI,
         prop_race_alone_asian,
         prop_race_alone_AI_AK,
         prop_race_alone_black,
         prop_minority)

# hispanic/latino
B03001_clean <-  B03001 %>%
  mutate(prop_hispanic_latino = `Estimate!!Total:!!Hispanic or Latino:`/`Estimate!!Total:`) %>%
  select(ZCTA,
         prop_hispanic_latino )

# nativity and citizenship status
B05001_clean <- B05001 %>%
  mutate(total=`Estimate!!Total:`,
         prop_not_citizen = `Estimate!!Total:!!Not a U.S. citizen`/total,
         prop_citizen = 1-prop_not_citizen) %>%
  select(ZCTA,
         prop_citizen)

# veteran status
B21001_clean <- B21001 %>%
  mutate(prop_veteran = `Estimate!!Total:!!Veteran`/`Estimate!!Total:`) %>%
  select(ZCTA,
         prop_veteran)

# household type
B11001_clean <- B11001 %>%
  mutate(prop_family_HH = `Estimate!!Total:!!Family households:`/`Estimate!!Total:`,
         prop_nonfamily_HH = `Estimate!!Total:!!Nonfamily households:`/`Estimate!!Total:`) %>%
  select(ZCTA,
         prop_family_HH)

# monthly housing costs
B25104_clean <- B25104 %>%
  mutate(prop_housing_no_cash_rent = `Estimate!!Total:!!No cash rent`/`Estimate!!Total:`,
         prop_housing_less_than_500 = (`Estimate!!Total:!!Less than $100`+
                                         `Estimate!!Total:!!$100 to $199`+
                                         `Estimate!!Total:!!$200 to $299`+
                                         `Estimate!!Total:!!$300 to $399`+
                                         `Estimate!!Total:!!$400 to $499`)/`Estimate!!Total:`,
         prop_housing_between_500_and_999 = (`Estimate!!Total:!!$500 to $599`+
                                               `Estimate!!Total:!!$600 to $699`+
                                               `Estimate!!Total:!!$700 to $799`+
                                               `Estimate!!Total:!!$800 to $899`+
                                               `Estimate!!Total:!!$900 to $999`)/`Estimate!!Total:`,
         prop_housing_between_1000_and_1500 = `Estimate!!Total:!!$1,000 to $1,499`/`Estimate!!Total:`,
         prop_housing_between_1500_and_2000 = `Estimate!!Total:!!$1,500 to $1,999`/`Estimate!!Total:`,
         prop_housing_between_2000_and_2500 = `Estimate!!Total:!!$2,000 to $2,499`/`Estimate!!Total:`,
         prop_housing_between_2500_and_3000 = `Estimate!!Total:!!$2,500 to $2,999`/`Estimate!!Total:`,
         prop_housing_greater_than_3000 = `Estimate!!Total:!!$3,000 or more`/`Estimate!!Total:`) %>%
  select(ZCTA,
         prop_housing_no_cash_rent,
         prop_housing_less_than_500,
         prop_housing_between_500_and_999,
         prop_housing_between_1000_and_1500,
         prop_housing_between_1500_and_2000,
         prop_housing_between_2000_and_2500,
         prop_housing_between_2500_and_3000,
         prop_housing_greater_than_3000)

# employment status 
B23001_clean <- B23001 %>%
  mutate(total_armed_forces = rowSums(B23001[,colnames(B23001)[grepl("in armed forces", colnames(B23001), ignore.case = TRUE)]]),
         total_employed = rowSums(B23001[,colnames(B23001)[grepl("Employed", colnames(B23001), ignore.case = FALSE)]]),
         total_unemployed = rowSums(B23001[,colnames(B23001)[grepl("unemployed", colnames(B23001), ignore.case = TRUE)]]),
         total_in_labor_force = total_employed + total_unemployed,
         total_not_in_labor_force = rowSums(B23001[,colnames(B23001)[grepl("not in labor force", colnames(B23001), ignore.case = TRUE)]])) %>%
  mutate(prop_armed_forces = total_armed_forces/`Estimate!!Total:`,
         prop_employed = total_employed/`Estimate!!Total:`,
         prop_unemployed = total_unemployed/`Estimate!!Total:`,
         prop_not_in_labor_force = total_not_in_labor_force/`Estimate!!Total:`) %>%
  select(ZCTA,
         prop_armed_forces,
         prop_employed,
         prop_unemployed,
         prop_not_in_labor_force)

# poverty status
B17001_clean <- B17001 %>%
  mutate(prop_at_or_above_poverty_level = `Estimate!!Total:!!Income in the past 12 months at or above poverty level:`/`Estimate!!Total:`,
         prop_below_poverty_level = `Estimate!!Total:!!Income in the past 12 months below poverty level:`/`Estimate!!Total:`) %>%
  select(ZCTA,
         prop_below_poverty_level)

# ratio of income to poverty level in past 12 months
B17024_clean <- B17024 %>%
  mutate(total_poverty_ratio_5.00_and_over = rowSums(B17024[,colnames(B17024)[grepl("5.00 and over", colnames(B17024), ignore.case = TRUE)]]),
         total_poverty_ratio_4.00_to_4.99 =  rowSums(B17024[,colnames(B17024)[grepl("4.00 to 4.99", colnames(B17024), ignore.case = TRUE)]]),
         total_poverty_ratio_3.00_to_3.99 =  rowSums(B17024[,colnames(B17024)[grepl("3.00 to 3.99", colnames(B17024), ignore.case = TRUE)]]),
         total_poverty_ratio_2.00_to_2.99 =  rowSums(B17024[,colnames(B17024)[grepl("2.00 to 2.99", colnames(B17024), ignore.case = TRUE)]]),
         total_poverty_ratio_1.85_to_1.99 =  rowSums(B17024[,colnames(B17024)[grepl("1.85 to 1.99", colnames(B17024), ignore.case = TRUE)]]),
         total_poverty_ratio_1.75_to_1.84 =  rowSums(B17024[,colnames(B17024)[grepl("1.75 to 1.84", colnames(B17024), ignore.case = TRUE)]]),
         total_poverty_ratio_1.50_to_1.74 =  rowSums(B17024[,colnames(B17024)[grepl("1.50 to 1.74", colnames(B17024), ignore.case = TRUE)]]),
         total_poverty_ratio_1.25_to_1.49 =  rowSums(B17024[,colnames(B17024)[grepl("1.25 to 1.49", colnames(B17024), ignore.case = TRUE)]]),
         total_poverty_ratio_1.00_to_1.24 =  rowSums(B17024[,colnames(B17024)[grepl("1.00 to 1.24", colnames(B17024), ignore.case = TRUE)]]),
         total_poverty_ratio_0.75_to_0.99 =  rowSums(B17024[,colnames(B17024)[grepl(".75 to .99", colnames(B17024), ignore.case = TRUE)]]),
         total_poverty_ratio_0.50_to_0.74 =  rowSums(B17024[,colnames(B17024)[grepl(".50 to .74", colnames(B17024), ignore.case = TRUE)]]),
         total_poverty_ratio_under_0.50 =  rowSums(B17024[,colnames(B17024)[grepl("under .50", colnames(B17024), ignore.case = TRUE)]])
  ) %>%
  mutate(prop_poverty_ratio_5.00_and_over = total_poverty_ratio_5.00_and_over / `Estimate!!Total:`,
         prop_poverty_ratio_4.00_to_4.99 = total_poverty_ratio_4.00_to_4.99 / `Estimate!!Total:`,
         prop_poverty_ratio_3.00_to_3.99 = total_poverty_ratio_3.00_to_3.99 / `Estimate!!Total:`,
         prop_poverty_ratio_2.00_to_2.99 = total_poverty_ratio_2.00_to_2.99 / `Estimate!!Total:`,
         prop_poverty_ratio_1.85_to_1.99 = total_poverty_ratio_1.85_to_1.99 / `Estimate!!Total:`,
         prop_poverty_ratio_1.75_to_1.84 = total_poverty_ratio_1.75_to_1.84 / `Estimate!!Total:`,
         prop_poverty_ratio_1.50_to_1.74 = total_poverty_ratio_1.50_to_1.74 / `Estimate!!Total:`,
         prop_poverty_ratio_1.25_to_1.49 = total_poverty_ratio_1.25_to_1.49 / `Estimate!!Total:`,
         prop_poverty_ratio_1.00_to_1.24 = total_poverty_ratio_1.00_to_1.24 / `Estimate!!Total:`,
         prop_poverty_ratio_0.75_to_0.99 = total_poverty_ratio_0.75_to_0.99 / `Estimate!!Total:`,
         prop_poverty_ratio_0.50_to_0.74 = total_poverty_ratio_0.50_to_0.74 / `Estimate!!Total:`,
         prop_poverty_ratio_under_0.50 = total_poverty_ratio_under_0.50 / `Estimate!!Total:`) %>%
  mutate(prop_poverty_ratio_under_1.50 = prop_poverty_ratio_1.25_to_1.49 +
                                         prop_poverty_ratio_1.00_to_1.24 +
                                         prop_poverty_ratio_0.75_to_0.99 +
                                         prop_poverty_ratio_0.50_to_0.74 +
                                         prop_poverty_ratio_under_0.50) %>%
  select(ZCTA,
         prop_poverty_ratio_under_1.50)

# median household income in the past 12 months (2021 inflation adjusted dollars)
B19013_clean <- B19013 %>%
  mutate(median_HH_income = `Estimate!!Median household income in the past 12 months (in 2023 inflation-adjusted dollars)`,
         median_HH_income = ifelse(median_HH_income < 0, NA, median_HH_income))%>%
  select(ZCTA,
         median_HH_income)

# Gini index of income inequality
B19083_clean <- B19083 %>%
  mutate(gini_index = `Estimate!!Gini Index`,
         gini_index = ifelse(gini_index < 0, NA, gini_index)) %>%
  select(ZCTA,
         gini_index)

# received food stamps/SNAP in the past 12 months for households
B22003_clean <- B22003 %>%
  mutate(prop_food_stamps = `Estimate!!Total:!!Household received Food Stamps/SNAP in the past 12 months:`/`Estimate!!Total:`,
         prop_no_food_stamps= `Estimate!!Total:!!Household did not receive Food Stamps/SNAP in the past 12 months:`/`Estimate!!Total:`) %>%
  select(ZCTA,
         prop_food_stamps)

# educational attainment for 18+
B15001_clean <- B15001 %>%
  mutate(total_no_high_school = rowSums(B15001[,colnames(B15001)[grepl("less than 9th grade", colnames(B15001), ignore.case = TRUE)]]),
         total_some_high_school = rowSums(B15001[,colnames(B15001)[grepl("9th to 12th grade", colnames(B15001), ignore.case = TRUE)]]),
         total_high_school_grad = rowSums(B15001[,colnames(B15001)[grepl("high school", colnames(B15001), ignore.case = TRUE)]]),
         total_some_college = rowSums(B15001[,colnames(B15001)[grepl("some college", colnames(B15001), ignore.case = TRUE)]]),
         total_associates_degree = rowSums(B15001[,colnames(B15001)[grepl("associate's degree", colnames(B15001), ignore.case = TRUE)]]),
         total_bachelors_degree = rowSums(B15001[,colnames(B15001)[grepl("bachelor's degree", colnames(B15001), ignore.case = TRUE)]]),
         total_graduate_degree_or_above = rowSums(B15001[,colnames(B15001)[grepl("graduate or professional", colnames(B15001), ignore.case = TRUE)]])) %>%
  mutate(prop_no_high_school = total_no_high_school/`Estimate!!Total:`,
         prop_some_high_school = total_some_high_school/`Estimate!!Total:`,
         prop_high_school_grad = total_high_school_grad/`Estimate!!Total:`,
         prop_some_college = total_some_college/`Estimate!!Total:`,
         prop_associates_degree = total_associates_degree/`Estimate!!Total:`,
         prop_bachelors_degree = total_bachelors_degree/`Estimate!!Total:`,
         prop_graduate_degree_or_above = total_graduate_degree_or_above/`Estimate!!Total:`) %>%
  mutate(prop_no_HS_diploma = prop_no_high_school + prop_some_high_school) %>%
  select(ZCTA,
         prop_no_HS_diploma,
         prop_high_school_grad,
         prop_some_college,
         prop_associates_degree,
         prop_bachelors_degree,
         prop_graduate_degree_or_above)
  
# means of transportation to work
B08006_clean <- B08006 %>%
  mutate(prop_car_truck_van = `Estimate!!Total:!!Car, truck, or van:`/`Estimate!!Total:`,
         prop_public_trans = `Estimate!!Total:!!Public transportation (excluding taxicab):`/`Estimate!!Total:`,
         prop_taxi_motorcycle_other = `Estimate!!Total:!!Taxicab, motorcycle, or other means`/`Estimate!!Total:`,
         prop_bicycle = `Estimate!!Total:!!Bicycle`/`Estimate!!Total:`,
         prop_walk = `Estimate!!Total:!!Walked`/`Estimate!!Total:`,
         prop_WFH = `Estimate!!Total:!!Worked from home`/`Estimate!!Total:`) %>%
  select(ZCTA,
         prop_car_truck_van,
         prop_public_trans,
         prop_taxi_motorcycle_other,
         prop_bicycle,
         prop_walk,
         prop_WFH)

# health insurance coverage status
B27001_clean <- B27001 %>%
  mutate(total_health_insurance = rowSums(B27001[,colnames(B27001)[grepl("with health insurance", colnames(B27001), ignore.case = TRUE)]]),
         total_no_health_insurance = rowSums(B27001[,colnames(B27001)[grepl("no health insurance", colnames(B27001), ignore.case = TRUE)]])) %>%
  mutate(prop_health_insurance = total_health_insurance/`Estimate!!Total:`,
         prop_no_health_insurance = total_no_health_insurance/`Estimate!!Total:`) %>%
  select(ZCTA,
         prop_health_insurance)

# disability status
B18101_clean <- B18101 %>%
  mutate(total_disability = rowSums(B18101[,colnames(B18101)[grepl("no disability", colnames(B18101), ignore.case = TRUE)]]),
         total_no_disability = rowSums(B18101[,colnames(B18101)[grepl("with a disability", colnames(B18101), ignore.case = TRUE)]])) %>%
  mutate(prop_disability = total_disability / `Estimate!!Total:`,
         prop_no_disability = total_no_disability / `Estimate!!Total:`) %>%
  select(ZCTA,
         prop_disability)

# limited English
B16005_clean <- B16005 %>%
  mutate(prop_limited_eng = (`Estimate!!Total:!!Foreign born:!!Speak other Indo-European languages:!!Speak English "not at all"`+
                               `Estimate!!Total:!!Foreign born:!!Speak Asian and Pacific Island languages:!!Speak English "not at all"` +
                               `Estimate!!Total:!!Native:!!Speak other Indo-European languages:!!Speak English "not at all"` +
                               `Estimate!!Total:!!Native:!!Speak Asian and Pacific Island languages:!!Speak English "not at all"` +
                               `Estimate!!Total:!!Foreign born:!!Speak other languages:!!Speak English "not at all"` +
                               `Estimate!!Total:!!Native:!!Speak other languages:!!Speak English "not at all"` +
                               `Estimate!!Total:!!Foreign born:!!Speak other Indo-European languages:!!Speak English "not well"` +
                               `Estimate!!Total:!!Foreign born:!!Speak Spanish:!!Speak English "not at all"` +
                               `Estimate!!Total:!!Foreign born:!!Speak Asian and Pacific Island languages:!!Speak English "not well"` +
                               `Estimate!!Total:!!Native:!!Speak other Indo-European languages:!!Speak English "not well"` +
                               `Estimate!!Total:!!Native:!!Speak Spanish:!!Speak English "not at all"` +
                               `Estimate!!Total:!!Native:!!Speak Asian and Pacific Island languages:!!Speak English "not well"` +
                               `Estimate!!Total:!!Foreign born:!!Speak Spanish:!!Speak English "not well"`+
                               `Estimate!!Total:!!Foreign born:!!Speak other languages:!!Speak English "not at all"` +
                               `Estimate!!Total:!!Native:!!Speak Spanish:!!Speak English "not well"` +
                               `Estimate!!Total:!!Native:!!Speak other languages:!!Speak English "not well"`) / `Estimate!!Total:`) %>%
  select(ZCTA,
         prop_limited_eng)

# no internet
B28011_clean <- B28011 %>%
  mutate(prop_no_internet = `Estimate!!Total:!!No Internet access`/`Estimate!!Total:`) %>%
  select(ZCTA,
         prop_no_internet)

# combine cleaned data
acs <- B01001_clean %>%
  left_join(B05001_clean, by="ZCTA") %>%
  left_join(B02001_clean, by="ZCTA") %>%
  left_join(B03001_clean, by="ZCTA") %>%
  left_join(B21001_clean, by="ZCTA") %>%
  left_join(B11001_clean, by="ZCTA") %>%
  left_join(B25104_clean, by="ZCTA") %>%
  left_join(B23001_clean, by="ZCTA") %>%
  left_join(B17001_clean, by="ZCTA") %>%
  left_join(B17024_clean, by="ZCTA") %>%
  left_join(B22003_clean, by="ZCTA") %>%
  left_join(B19013_clean, by="ZCTA") %>%
  left_join(B19083_clean, by="ZCTA") %>%
  left_join(B15001_clean, by="ZCTA") %>%
  left_join(B08006_clean, by="ZCTA") %>%
  left_join(B27001_clean, by="ZCTA") %>%
  left_join(B18101_clean, by="ZCTA") %>%
  left_join(B16005_clean, by="ZCTA") %>%
  left_join(B28011_clean, by="ZCTA") %>%
  mutate(ZCTA = as.numeric(ZCTA))

# add ACS_ to column names
colnames(acs)[-1] <- paste0("ACS_", colnames(acs)[-1])
```

##################################################################
############### Local Data for Better Health (CDC) ###############
##################################################################

# Import and clean CDC data
## change to most recent file
```{r import and clean CDC data}
# read in CDC disease burden file
cdc_file <- read.csv("F:\\Data_Science\\SDoH\\sdoh 2025 update\\PLACES__Local_Data_for_Better_Health__ZCTA_Data_2024_release_20250402.csv")

# variable descriptions
cdc_description <- cdc_file %>%
  select(Measure,
         MeasureId) %>%
  distinct()
  
# only select crude prevalence for each disease category at ZCTA level
cdc <- cdc_file %>%
  filter(Data_Value_Type=="Crude prevalence") %>%
  mutate(ZCTA=LocationName,
         prevalence=Data_Value,
         disease_category=MeasureId) %>%
  select(ZCTA,
         prevalence,
         disease_category) %>%

  # format as a "wide" table
  tidyr::pivot_wider(names_from = disease_category,
              values_from = prevalence) %>%
  mutate(ZCTA=as.numeric(ZCTA)) %>%
  select(all_of(sort(names(.)))) %>%
  relocate(ZCTA)

# format as percentages
cdc[colnames(cdc)[-1]] <- cdc[colnames(cdc)[-1]] / 100

# add CDC_ to column names
colnames(cdc)[-1] <- paste0("CDC_", colnames(cdc)[-1])
```

####################################################################
############### CDC Social Vulnerability Index (SVI) ###############
####################################################################

# Import and clean SVI data
## change to most recent file
## now SVI has ZCTA-level data so do not have to use tract-level data anymore, no weighting needed
```{r import and clean SVI data}
svi_file <- read.csv("F:\\Data_Science\\SDoH\\sdoh 2025 update\\SVI_2022_US_ZCTA.csv")

svi <- svi_file %>%
  mutate(ZCTA = as.numeric(gsub("ZCTA5 ", "", LOCATION))) %>%
  select(ZCTA,
         colnames(svi_file)[grepl("EP_", colnames(svi_file), ignore.case = TRUE)]) %>%
  # change -999's to NA
  mutate(EP_UNEMP = ifelse(EP_UNEMP < 0, NA, EP_UNEMP),
         EP_NOHSDP = ifelse(EP_NOHSDP < 0, NA, EP_NOHSDP),
         EP_UNINSUR = ifelse(EP_UNINSUR < 0, NA, EP_UNINSUR),
         EP_AGE65 = ifelse(EP_AGE65 < 0, NA, EP_AGE65),
         EP_DISABL = ifelse(EP_DISABL < 0, NA, EP_DISABL),
         EP_MOBILE = ifelse(EP_MOBILE < 0, NA, EP_MOBILE),
         EP_NOVEH = ifelse(EP_NOVEH < 0, NA, EP_NOVEH),
         EP_AFAM = ifelse(EP_AFAM < 0, NA, EP_AFAM),
         EP_HISP = ifelse(EP_HISP < 0, NA, EP_HISP),
         EP_ASIAN = ifelse(EP_ASIAN < 0, NA, EP_ASIAN),
         EP_AIAN = ifelse(EP_AIAN < 0, NA, EP_AIAN),
         EP_NHPI = ifelse(EP_NHPI < 0, NA, EP_NHPI),
         EP_TWOMORE = ifelse(EP_TWOMORE < 0, NA, EP_TWOMORE),
         EP_OTHERRACE = ifelse(EP_OTHERRACE < 0, NA, EP_OTHERRACE))

svi_EP_cols <- colnames(svi %>% select(contains("EP")))
svi[svi_EP_cols] <- svi[svi_EP_cols] / 100

# add SVI_ to column names
colnames(svi)[-1] <- paste0("SVI_", colnames(svi)[-1])
```

#####################################################################
############### Create Zip Code-Level Aggregate Table ###############
#####################################################################

## change to most recent file
## include region as a variable now
```{r create zip code table}
### map ZCTA to zip code to state
# import zcta to zip crosswalk from UDS Mapper (2022 version)
zcta_to_zip_file <- readxl::read_excel("F:\\Data_Science\\SDoH\\sdoh 2025 update\\ZIPCodetoZCTACrosswalk2022UDS.xlsx")
zcta_to_zip <- zcta_to_zip_file %>%
  mutate(ZCTA = as.numeric(zcta)) %>%
  select(STATE,
         ZCTA,
         ZIP_CODE)

### map state to region
state_to_region_file <- readxl::read_excel("F:\\Data_Science\\SDoH\\Data\\sdoh 2024\\state to region mapping 20250121.xlsx")

table_zip <- acs %>%
  
  left_join(cdc, by="ZCTA") %>%
  
  left_join(svi, by="ZCTA") %>%
  
  # add state and zip code
  left_join(zcta_to_zip, by="ZCTA") %>%
  
  # add region
  left_join(state_to_region_file, by="STATE") %>%
  
  # reorder variables
  relocate(REGION,
           STATE,
           ZIP_CODE) %>%
  select(-ZCTA) %>%
  filter(!is.na(ZIP_CODE)) 

# save as RDS
# saveRDS(table_zip, file = "F:\\Data_Science\\SDoH\\sdoh 2025 update\\SDoH_zip_table 20250402.RDS")
```

##########################################################################################################
########################################## County Level Table ############################################
##########################################################################################################

###############################################################
######## NCHS Urban-Rural Classification Scheme (NCHS) ########
###############################################################

# Import NCHS data
```{r import and clean NCHS data}
nchs_file <- read.csv("F:\\Data_Science\\SDoH\\sdoh 2025 update\\NCHS_2023_20250402.csv")
nchs <- nchs_file %>%
  mutate(FIPS = as.character(FIPS.code),
         NCHS_urban_rural_score = as.numeric(substr(X2023.Code, 1, 1))) %>%
  select(FIPS,
         NCHS_urban_rural_score) %>%
  mutate(FIPS = as.character(FIPS))
```

###############################################################
############### American Community Survey (ACS) ###############
###############################################################

# Import and clean ACS data
```{r pull ACS data from Census Bureau website and clean}
# dictionary
dictionary_county <- as.data.frame(rbind(
  c("B01001", "population by sex and age"),
  c("B02001", "race"),
  c("B03001", "hispanic/latino"),
  c("B05001", "citizenship"),
  c("B21001", "veteran status"),
  c("B11001", "household type"),
  c("B25104", "monthly housing costs"),
  c("B23001", "employment status"),
  c("B17001", "poverty status"),
  c("B17024", "ratio of income to poverty level"),
  c("B19013", "median household income"),
  c("B19083", "Gini index of income inequality"),
  c("B22003", "food stamps/SNAP"),
  c("B15001", "educational attainment for 18+"),
  c("B08006", "means of transportation to work"),
  c("B27001", "health insurance coverage status"),
  c("B18101", "disability status"),
  c("B16005", "limited English"),
  c("B28011", "no internet")
) )
colnames(dictionary_county) <- c("table", "topic")

# establish Census API connection
Sys.setenv(CENSUS_KEY= "aee7b281866941d1d77418ed939221a02d29ea97")
readRenviron("~/.Renviron")
Sys.getenv("CENSUS_KEY")

# build function to execute census api call
censusapicall_county <- function(x){
  
  # get variable names
  variables <- listCensusMetadata(
    name="acs/acs5",
    vintage="2023",
    type="variables",
    group=x) %>%
    filter(grepl("EA",name)==FALSE,
           grepl("MA",name)==FALSE,
           grepl("M",name)==FALSE)
  
  # select variable names and descriptions
  vars <- variables$name
  cols <- variables$label
  
  # pull selected variables
  data <- censusapi::getCensus(
    name="acs/acs5",
    vintage="2023",
    vars=c("NAME", vars),
    region="county:*") %>%
    mutate(FIPS=as.character(as.numeric(paste0(state, county)))) %>%
    relocate(FIPS) %>%
    select(-c(state, county, NAME))
    
  
  # adding column names to data
  colnames(data) <- c("FIPS", cols)
  
  # create table for each topic
  assign(paste0(x, "_county"), data, envir = .GlobalEnv)
  
}

# use censusapicall function to pull selected tables
api_calls_county <- sapply(dictionary_county$table, censusapicall_county)

# population by sex and age
B01001_clean_county <- B01001_county %>%
  mutate(total_pop = `Estimate!!Total:`,
         prop_male = `Estimate!!Total:!!Male:`/`Estimate!!Total:`,
         prop_female = `Estimate!!Total:!!Female:`/`Estimate!!Total:`,
         total_age_under_18 = (  rowSums(B01001_county[,colnames(B01001_county)[grepl("under 5 years", colnames(B01001_county), ignore.case = TRUE)]]) +
                                   rowSums(B01001_county[,colnames(B01001_county)[grepl("5 to 9 years", colnames(B01001_county), ignore.case = TRUE)]]) +
                                   rowSums(B01001_county[,colnames(B01001_county)[grepl("10 to 14 years", colnames(B01001_county), ignore.case = TRUE)]]) +
                                   rowSums(B01001_county[,colnames(B01001_county)[grepl("15 to 17 years", colnames(B01001_county), ignore.case = TRUE)]]) ) ,
         total_age_18_to_64 = (  rowSums(B01001_county[,colnames(B01001_county)[grepl("18 and 19 years", colnames(B01001_county), ignore.case = TRUE)]]) +
                                   rowSums(B01001_county[,colnames(B01001_county)[grepl("20 years", colnames(B01001_county), ignore.case = TRUE)]]) +
                                   rowSums(B01001_county[,colnames(B01001_county)[grepl("21 years", colnames(B01001_county), ignore.case = TRUE)]]) +
                                   rowSums(B01001_county[,colnames(B01001_county)[grepl("22 to 24 years", colnames(B01001_county), ignore.case = TRUE)]]) +
                                   rowSums(B01001_county[,colnames(B01001_county)[grepl("25 to 29 years", colnames(B01001_county), ignore.case = TRUE)]]) +
                                   rowSums(B01001_county[,colnames(B01001_county)[grepl("30 to 34 years", colnames(B01001_county), ignore.case = TRUE)]]) +
                                   rowSums(B01001_county[,colnames(B01001_county)[grepl("35 to 39 years", colnames(B01001_county), ignore.case = TRUE)]]) +
                                   rowSums(B01001_county[,colnames(B01001_county)[grepl("40 to 44 years", colnames(B01001_county), ignore.case = TRUE)]]) +
                                   rowSums(B01001_county[,colnames(B01001_county)[grepl("45 to 49 years", colnames(B01001_county), ignore.case = TRUE)]]) +
                                   rowSums(B01001_county[,colnames(B01001_county)[grepl("50 to 54 years", colnames(B01001_county), ignore.case = TRUE)]]) +
                                   rowSums(B01001_county[,colnames(B01001_county)[grepl("55 to 59 years", colnames(B01001_county), ignore.case = TRUE)]]) +
                                   rowSums(B01001_county[,colnames(B01001_county)[grepl("60 and 61 years", colnames(B01001_county), ignore.case = TRUE)]]) +
                                   rowSums(B01001_county[,colnames(B01001_county)[grepl("62 to 64 years", colnames(B01001_county), ignore.case = TRUE)]]) ) ,
         total_age_over_64 = (  rowSums(B01001_county[,colnames(B01001_county)[grepl("65 and 66 years", colnames(B01001_county), ignore.case = TRUE)]]) +
                                  rowSums(B01001_county[,colnames(B01001_county)[grepl("67 to 69 years", colnames(B01001_county), ignore.case = TRUE)]]) +
                                  rowSums(B01001_county[,colnames(B01001_county)[grepl("70 to 74 years", colnames(B01001_county), ignore.case = TRUE)]]) +
                                  rowSums(B01001_county[,colnames(B01001_county)[grepl("75 to 79 years", colnames(B01001_county), ignore.case = TRUE)]]) +
                                  rowSums(B01001_county[,colnames(B01001_county)[grepl("80 to 84 years", colnames(B01001_county), ignore.case = TRUE)]]) +
                                  rowSums(B01001_county[,colnames(B01001_county)[grepl("85 years and over", colnames(B01001_county), ignore.case = TRUE)]]) ),
         prop_age_under_18 = total_age_under_18/`Estimate!!Total:`,
         prop_age_18_to_64 = total_age_18_to_64/`Estimate!!Total:`,
         prop_age_over_64 = total_age_over_64/`Estimate!!Total:` ) %>%
  select(FIPS,
         #total_pop,
         # prop_male,
         prop_female,
         prop_age_under_18,
         prop_age_18_to_64,
         prop_age_over_64)

# race
B02001_clean_county <- B02001_county %>%
  mutate(total_race = `Estimate!!Total:`,
         prop_race_two_or_more = `Estimate!!Total:!!Two or More Races:`/total_race,
         prop_race_alone_other = `Estimate!!Total:!!Some Other Race alone`/total_race,
         prop_race_alone_white = `Estimate!!Total:!!White alone`/total_race,
         prop_race_alone_HI_PI = `Estimate!!Total:!!Native Hawaiian and Other Pacific Islander alone`/total_race,
         prop_race_alone_asian = `Estimate!!Total:!!Asian alone`/total_race,
         prop_race_alone_AI_AK = `Estimate!!Total:!!American Indian and Alaska Native alone`/total_race,
         prop_race_alone_black = `Estimate!!Total:!!Black or African American alone`/total_race,
         prop_minority = (prop_race_two_or_more +
                             prop_race_alone_other +
                             prop_race_alone_HI_PI +
                             prop_race_alone_asian +
                             prop_race_alone_AI_AK +
                             prop_race_alone_black) ) %>%
  select(FIPS,
         prop_race_two_or_more,
         prop_race_alone_other,
         prop_race_alone_white,
         prop_race_alone_HI_PI,
         prop_race_alone_asian,
         prop_race_alone_AI_AK,
         prop_race_alone_black,
         prop_minority)

# hispanic/latino
B03001_clean_county <-  B03001_county %>%
  mutate(prop_hispanic_latino = `Estimate!!Total:!!Hispanic or Latino:`/`Estimate!!Total:`,
         prop_non_hispanic_latino = `Estimate!!Total:!!Not Hispanic or Latino`/`Estimate!!Total:`) %>%
  select(FIPS,
         prop_hispanic_latino)
         # prop_non_hispanic_latino)

# nativity and citizenship status
B05001_clean_county <- B05001_county %>%
  mutate(total=`Estimate!!Total:`,
         prop_not_citizen = `Estimate!!Total:!!Not a U.S. citizen`/total,
         prop_citizen = 1-prop_not_citizen) %>%
  select(FIPS,
         prop_citizen)
         # prop_not_citizen)

# veteran status
B21001_clean_county <- B21001_county %>%
  mutate(prop_veteran = `Estimate!!Total:!!Veteran`/`Estimate!!Total:`,
         prop_nonveteran =`Estimate!!Total:!!Nonveteran`/`Estimate!!Total:`) %>%
  select(FIPS,
         prop_veteran)
         # prop_nonveteran)

# household type
B11001_clean_county <- B11001_county %>%
  mutate(prop_family_HH = `Estimate!!Total:!!Family households:`/`Estimate!!Total:`,
         prop_nonfamily_HH = `Estimate!!Total:!!Nonfamily households:`/`Estimate!!Total:`) %>%
  select(FIPS,
         prop_family_HH)
         # prop_nonfamily_HH)

# monthly housing costs
B25104_clean_county <- B25104_county %>%
  mutate(prop_housing_no_cash_rent = `Estimate!!Total:!!No cash rent`/`Estimate!!Total:`,
         prop_housing_less_than_500 = (`Estimate!!Total:!!Less than $100`+
                                         `Estimate!!Total:!!$100 to $199`+
                                         `Estimate!!Total:!!$200 to $299`+
                                         `Estimate!!Total:!!$300 to $399`+
                                         `Estimate!!Total:!!$400 to $499`)/`Estimate!!Total:`,
         prop_housing_between_500_and_999 = (`Estimate!!Total:!!$500 to $599`+
                                               `Estimate!!Total:!!$600 to $699`+
                                               `Estimate!!Total:!!$700 to $799`+
                                               `Estimate!!Total:!!$800 to $899`+
                                               `Estimate!!Total:!!$900 to $999`)/`Estimate!!Total:`,
         prop_housing_between_1000_and_1500 = `Estimate!!Total:!!$1,000 to $1,499`/`Estimate!!Total:`,
         prop_housing_between_1500_and_2000 = `Estimate!!Total:!!$1,500 to $1,999`/`Estimate!!Total:`,
         prop_housing_between_2000_and_2500 = `Estimate!!Total:!!$2,000 to $2,499`/`Estimate!!Total:`,
         prop_housing_between_2500_and_3000 = `Estimate!!Total:!!$2,500 to $2,999`/`Estimate!!Total:`,
         prop_housing_greater_than_3000 = `Estimate!!Total:!!$3,000 or more`/`Estimate!!Total:`) %>%
  select(FIPS,
         prop_housing_no_cash_rent,
         prop_housing_less_than_500,
         prop_housing_between_500_and_999,
         prop_housing_between_1000_and_1500,
         prop_housing_between_1500_and_2000,
         prop_housing_between_2000_and_2500,
         prop_housing_between_2500_and_3000,
         prop_housing_greater_than_3000)

# employment status 
B23001_clean_county <- B23001_county %>%
  mutate(total_armed_forces = rowSums(B23001_county[,colnames(B23001_county)[grepl("in armed forces", colnames(B23001_county), ignore.case = TRUE)]]),
         total_employed = rowSums(B23001_county[,colnames(B23001_county)[grepl("Employed", colnames(B23001_county), ignore.case = FALSE)]]),
         total_unemployed = rowSums(B23001_county[,colnames(B23001_county)[grepl("unemployed", colnames(B23001_county), ignore.case = TRUE)]]),
         total_in_labor_force = total_employed + total_unemployed,
         total_not_in_labor_force = rowSums(B23001_county[,colnames(B23001_county)[grepl("not in labor force", colnames(B23001_county), ignore.case = TRUE)]])) %>%
  mutate(prop_armed_forces = total_armed_forces/`Estimate!!Total:`,
         prop_employed = total_employed/`Estimate!!Total:`,
         prop_unemployed = total_unemployed/`Estimate!!Total:`,
         prop_not_in_labor_force = total_not_in_labor_force/`Estimate!!Total:`) %>%
  select(FIPS,
         prop_armed_forces,
         prop_employed,
         prop_unemployed,
         prop_not_in_labor_force)

# poverty status
B17001_clean_county <- B17001_county %>%
  mutate(prop_at_or_above_poverty_level = `Estimate!!Total:!!Income in the past 12 months at or above poverty level:`/`Estimate!!Total:`,
         prop_below_poverty_level = `Estimate!!Total:!!Income in the past 12 months below poverty level:`/`Estimate!!Total:`) %>%
  select(FIPS,
         # prop_at_or_above_poverty_level,
         prop_below_poverty_level)

# ratio of income to poverty level in past 12 months
B17024_clean_county <- B17024_county %>%
  mutate(total_poverty_ratio_5.00_and_over = rowSums(B17024_county[,colnames(B17024_county)[grepl("5.00 and over", colnames(B17024_county), ignore.case = TRUE)]]),
         total_poverty_ratio_4.00_to_4.99 =  rowSums(B17024_county[,colnames(B17024_county)[grepl("4.00 to 4.99", colnames(B17024_county), ignore.case = TRUE)]]),
         total_poverty_ratio_3.00_to_3.99 =  rowSums(B17024_county[,colnames(B17024_county)[grepl("3.00 to 3.99", colnames(B17024_county), ignore.case = TRUE)]]),
         total_poverty_ratio_2.00_to_2.99 =  rowSums(B17024_county[,colnames(B17024_county)[grepl("2.00 to 2.99", colnames(B17024_county), ignore.case = TRUE)]]),
         total_poverty_ratio_1.85_to_1.99 =  rowSums(B17024_county[,colnames(B17024_county)[grepl("1.85 to 1.99", colnames(B17024_county), ignore.case = TRUE)]]),
         total_poverty_ratio_1.75_to_1.84 =  rowSums(B17024_county[,colnames(B17024_county)[grepl("1.75 to 1.84", colnames(B17024_county), ignore.case = TRUE)]]),
         total_poverty_ratio_1.50_to_1.74 =  rowSums(B17024_county[,colnames(B17024_county)[grepl("1.50 to 1.74", colnames(B17024_county), ignore.case = TRUE)]]),
         total_poverty_ratio_1.25_to_1.49 =  rowSums(B17024_county[,colnames(B17024_county)[grepl("1.25 to 1.49", colnames(B17024_county), ignore.case = TRUE)]]),
         total_poverty_ratio_1.00_to_1.24 =  rowSums(B17024_county[,colnames(B17024_county)[grepl("1.00 to 1.24", colnames(B17024_county), ignore.case = TRUE)]]),
         total_poverty_ratio_0.75_to_0.99 =  rowSums(B17024_county[,colnames(B17024_county)[grepl(".75 to .99", colnames(B17024_county), ignore.case = TRUE)]]),
         total_poverty_ratio_0.50_to_0.74 =  rowSums(B17024_county[,colnames(B17024_county)[grepl(".50 to .74", colnames(B17024_county), ignore.case = TRUE)]]),
         total_poverty_ratio_under_0.50 =  rowSums(B17024_county[,colnames(B17024_county)[grepl("under .50", colnames(B17024_county), ignore.case = TRUE)]])
  ) %>%
  mutate(prop_poverty_ratio_5.00_and_over = total_poverty_ratio_5.00_and_over / `Estimate!!Total:`,
         prop_poverty_ratio_4.00_to_4.99 = total_poverty_ratio_4.00_to_4.99 / `Estimate!!Total:`,
         prop_poverty_ratio_3.00_to_3.99 = total_poverty_ratio_3.00_to_3.99 / `Estimate!!Total:`,
         prop_poverty_ratio_2.00_to_2.99 = total_poverty_ratio_2.00_to_2.99 / `Estimate!!Total:`,
         prop_poverty_ratio_1.85_to_1.99 = total_poverty_ratio_1.85_to_1.99 / `Estimate!!Total:`,
         prop_poverty_ratio_1.75_to_1.84 = total_poverty_ratio_1.75_to_1.84 / `Estimate!!Total:`,
         prop_poverty_ratio_1.50_to_1.74 = total_poverty_ratio_1.50_to_1.74 / `Estimate!!Total:`,
         prop_poverty_ratio_1.25_to_1.49 = total_poverty_ratio_1.25_to_1.49 / `Estimate!!Total:`,
         prop_poverty_ratio_1.00_to_1.24 = total_poverty_ratio_1.00_to_1.24 / `Estimate!!Total:`,
         prop_poverty_ratio_0.75_to_0.99 = total_poverty_ratio_0.75_to_0.99 / `Estimate!!Total:`,
         prop_poverty_ratio_0.50_to_0.74 = total_poverty_ratio_0.50_to_0.74 / `Estimate!!Total:`,
         prop_poverty_ratio_under_0.50 = total_poverty_ratio_under_0.50 / `Estimate!!Total:`) %>%
  mutate(prop_poverty_ratio_under_1.50 = prop_poverty_ratio_1.25_to_1.49 +
                                         prop_poverty_ratio_1.00_to_1.24 +
                                         prop_poverty_ratio_0.75_to_0.99 +
                                         prop_poverty_ratio_0.50_to_0.74 +
                                         prop_poverty_ratio_under_0.50) %>%
  select(FIPS,
         prop_poverty_ratio_under_1.50)

# median household income in the past 12 months (2023 inflation adjusted dollars)
B19013_clean_county <- B19013_county %>%
  mutate(median_HH_income = `Estimate!!Median household income in the past 12 months (in 2023 inflation-adjusted dollars)`,
         median_HH_income = ifelse(median_HH_income < 0, NA, median_HH_income))%>%
  select(FIPS,
         median_HH_income)

# Gini index of income inequality
B19083_clean_county <- B19083_county %>%
  mutate(gini_index = `Estimate!!Gini Index`) %>%
  select(FIPS,
         gini_index)

# received food stamps/SNAP in the past 12 months for households
B22003_clean_county <- B22003_county %>%
  mutate(prop_food_stamps = `Estimate!!Total:!!Household received Food Stamps/SNAP in the past 12 months:`/`Estimate!!Total:`,
         prop_no_food_stamps= `Estimate!!Total:!!Household did not receive Food Stamps/SNAP in the past 12 months:`/`Estimate!!Total:`) %>%
  select(FIPS,
         prop_food_stamps)
         # prop_no_food_stamps)

# educational attainment for 18+
B15001_clean_county <- B15001_county %>%
  mutate(total_no_high_school = rowSums(B15001_county[,colnames(B15001_county)[grepl("less than 9th grade", colnames(B15001_county), ignore.case = TRUE)]]),
         total_some_high_school = rowSums(B15001_county[,colnames(B15001_county)[grepl("9th to 12th grade", colnames(B15001_county), ignore.case = TRUE)]]),
         total_high_school_grad = rowSums(B15001_county[,colnames(B15001_county)[grepl("high school", colnames(B15001_county), ignore.case = TRUE)]]),
         total_some_college = rowSums(B15001_county[,colnames(B15001_county)[grepl("some college", colnames(B15001_county), ignore.case = TRUE)]]),
         total_associates_degree = rowSums(B15001_county[,colnames(B15001_county)[grepl("associate's degree", colnames(B15001_county), ignore.case = TRUE)]]),
         total_bachelors_degree = rowSums(B15001_county[,colnames(B15001_county)[grepl("bachelor's degree", colnames(B15001_county), ignore.case = TRUE)]]),
         total_graduate_degree_or_above = rowSums(B15001_county[,colnames(B15001_county)[grepl("graduate or professional", colnames(B15001_county), ignore.case = TRUE)]])) %>%
  mutate(prop_no_high_school = total_no_high_school/`Estimate!!Total:`,
         prop_some_high_school = total_some_high_school/`Estimate!!Total:`,
         prop_high_school_grad = total_high_school_grad/`Estimate!!Total:`,
         prop_some_college = total_some_college/`Estimate!!Total:`,
         prop_associates_degree = total_associates_degree/`Estimate!!Total:`,
         prop_bachelors_degree = total_bachelors_degree/`Estimate!!Total:`,
         prop_graduate_degree_or_above = total_graduate_degree_or_above/`Estimate!!Total:`) %>%
  mutate(prop_no_HS_diploma = prop_no_high_school + prop_some_high_school) %>%
  select(FIPS,
         prop_no_HS_diploma,
         prop_high_school_grad,
         prop_some_college,
         prop_associates_degree,
         prop_bachelors_degree,
         prop_graduate_degree_or_above)

# means of transportation to work
B08006_clean_county <- B08006_county %>%
  mutate(prop_car_truck_van = `Estimate!!Total:!!Car, truck, or van:`/`Estimate!!Total:`,
         prop_public_trans = `Estimate!!Total:!!Public transportation (excluding taxicab):`/`Estimate!!Total:`,
         prop_taxi_motorcycle_other = `Estimate!!Total:!!Taxicab, motorcycle, or other means`/`Estimate!!Total:`,
         prop_bicycle = `Estimate!!Total:!!Bicycle`/`Estimate!!Total:`,
         prop_walk = `Estimate!!Total:!!Walked`/`Estimate!!Total:`,
         prop_WFH = `Estimate!!Total:!!Worked from home`/`Estimate!!Total:`) %>%
  select(FIPS,
         prop_car_truck_van,
         prop_public_trans,
         prop_taxi_motorcycle_other,
         prop_bicycle,
         prop_walk,
         prop_WFH)

# health insurance coverage status
B27001_clean_county <- B27001_county %>%
  mutate(total_health_insurance = rowSums(B27001_county[,colnames(B27001_county)[grepl("with health insurance", colnames(B27001_county), ignore.case = TRUE)]]),
         total_no_health_insurance = rowSums(B27001_county[,colnames(B27001_county)[grepl("no health insurance", colnames(B27001_county), ignore.case = TRUE)]])) %>%
  mutate(prop_health_insurance = total_health_insurance/`Estimate!!Total:`,
         prop_no_health_insurance = total_no_health_insurance/`Estimate!!Total:`) %>%
  select(FIPS,
         prop_health_insurance)
         # prop_no_health_insurance)

# disability status
B18101_clean_county <- B18101_county %>%
  mutate(total_disability = rowSums(B18101_county[,colnames(B18101_county)[grepl("no disability", colnames(B18101_county), ignore.case = TRUE)]]),
         total_no_disability = rowSums(B18101_county[,colnames(B18101_county)[grepl("with a disability", colnames(B18101_county), ignore.case = TRUE)]])) %>%
  mutate(prop_disability = total_disability / `Estimate!!Total:`,
         prop_no_disability = total_no_disability / `Estimate!!Total:`) %>%
  select(FIPS,
         prop_disability)
         # prop_no_disability)

# limited English
B16005_clean_county <- B16005_county %>%
  mutate(prop_limited_eng = (`Estimate!!Total:!!Foreign born:!!Speak other Indo-European languages:!!Speak English "not at all"`+
                               `Estimate!!Total:!!Foreign born:!!Speak Asian and Pacific Island languages:!!Speak English "not at all"` +
                               `Estimate!!Total:!!Native:!!Speak other Indo-European languages:!!Speak English "not at all"` +
                               `Estimate!!Total:!!Native:!!Speak Asian and Pacific Island languages:!!Speak English "not at all"` +
                               `Estimate!!Total:!!Foreign born:!!Speak other languages:!!Speak English "not at all"` +
                               `Estimate!!Total:!!Native:!!Speak other languages:!!Speak English "not at all"` +
                               `Estimate!!Total:!!Foreign born:!!Speak other Indo-European languages:!!Speak English "not well"` +
                               `Estimate!!Total:!!Foreign born:!!Speak Spanish:!!Speak English "not at all"` +
                               `Estimate!!Total:!!Foreign born:!!Speak Asian and Pacific Island languages:!!Speak English "not well"` +
                               `Estimate!!Total:!!Native:!!Speak other Indo-European languages:!!Speak English "not well"` +
                               `Estimate!!Total:!!Native:!!Speak Spanish:!!Speak English "not at all"` +
                               `Estimate!!Total:!!Native:!!Speak Asian and Pacific Island languages:!!Speak English "not well"` +
                               `Estimate!!Total:!!Foreign born:!!Speak Spanish:!!Speak English "not well"`+
                               `Estimate!!Total:!!Foreign born:!!Speak other languages:!!Speak English "not at all"` +
                               `Estimate!!Total:!!Native:!!Speak Spanish:!!Speak English "not well"` +
                               `Estimate!!Total:!!Native:!!Speak other languages:!!Speak English "not well"`) / `Estimate!!Total:`) %>%
  select(FIPS,
         prop_limited_eng)

# no internet
B28011_clean_county <- B28011_county %>%
  mutate(prop_no_internet = `Estimate!!Total:!!No Internet access`/`Estimate!!Total:`) %>%
  select(FIPS,
         prop_no_internet)

# combine cleaned data
acs_county <- B01001_clean_county %>%
  left_join(B05001_clean_county, by="FIPS") %>%
  left_join(B02001_clean_county, by="FIPS") %>%
  left_join(B03001_clean_county, by="FIPS") %>%
  left_join(B21001_clean_county, by="FIPS") %>%
  left_join(B11001_clean_county, by="FIPS") %>%
  left_join(B25104_clean_county, by="FIPS") %>%
  left_join(B23001_clean_county, by="FIPS") %>%
  left_join(B17001_clean_county, by="FIPS") %>%
  left_join(B17024_clean_county, by="FIPS") %>%
  left_join(B22003_clean_county, by="FIPS") %>%
  left_join(B19013_clean_county, by="FIPS") %>%
  left_join(B19083_clean_county, by="FIPS") %>%
  left_join(B15001_clean_county, by="FIPS") %>%
  left_join(B08006_clean_county, by="FIPS") %>%
  left_join(B27001_clean_county, by="FIPS") %>%
  left_join(B18101_clean_county, by="FIPS") %>%
  left_join(B16005_clean_county, by="FIPS") %>%
  left_join(B28011_clean_county, by="FIPS")

# add ACS_ to column names
colnames(acs_county)[-1] <- paste0("ACS_", colnames(acs_county)[-1])
```

##################################################################
############### Local Data for Better Health (CDC) ###############
##################################################################

# Import and clean CDC data
```{r import and clean CDC data}
# read in CDC disease burden file
cdc_county_file <- read.csv("F:\\Data_Science\\SDoH\\sdoh 2025 update\\PLACES__Local_Data_for_Better_Health__County_Data_2024_release_20250402.csv")

# variable descriptions
cdc_county_description <- cdc_county_file %>%
  select(Measure,
         MeasureId) %>%
  distinct()

# only select crude prevalence for each disease category at county level
cdc_county <- cdc_county_file %>%
  filter(Data_Value_Type=="Crude prevalence") %>%
  mutate(FIPS=LocationID,
         prevalence=Data_Value,
         disease_category=MeasureId) %>%
  select(FIPS,
         prevalence,
         disease_category) %>%
  
  # format as a "wide" table
  tidyr::pivot_wider(names_from = disease_category,
                     values_from = prevalence) %>%
  mutate(FIPS=as.character(FIPS))

# format as percentages
cdc_county[colnames(cdc_county)[-1]] <- cdc_county[colnames(cdc_county)[-1]] / 100

# add CDC_ to column names
colnames(cdc_county)[-1] <- paste0("CDC_", colnames(cdc_county)[-1])
```

####################################################################
############### CDC Social Vulnerability Index (SVI) ###############
####################################################################

# Import and clean SVI data
```{r import and clean SVI data}
svi_county_file <- read.csv("F:\\Data_Science\\SDoH\\sdoh 2025 update\\SVI_2022_US_county.csv")

svi_county <- svi_county_file %>%
  select(STATE,
         COUNTY,
         FIPS,
         colnames(svi_file)[grepl("EP_", colnames(svi_file), ignore.case = TRUE)],
         colnames(svi_file)[grepl("EPL_", colnames(svi_file), ignore.case = TRUE)],
         colnames(svi_file)[grepl("F_", colnames(svi_file), ignore.case = TRUE)],
         colnames(svi_file)[grepl("SPL_", colnames(svi_file), ignore.case = TRUE)],
         colnames(svi_file)[grepl("RPL_", colnames(svi_file), ignore.case = TRUE)]) %>%
  mutate(FIPS = as.character(FIPS)) %>%
  
  # convert -999's to NA's
  mutate(F_POV150 = if_else(F_POV150 < 0, NA, F_POV150),
         F_UNEMP = if_else(F_POV150 < 0, NA, F_POV150),
         F_HBURD = if_else(F_POV150 < 0, NA, F_POV150),
         F_NOHSDP = if_else(F_POV150 < 0, NA, F_POV150),
         F_UNINSUR = if_else(F_POV150 < 0, NA, F_POV150),
         F_THEME1 = if_else(F_POV150 < 0, NA, F_POV150),
         F_AGE65 = if_else(F_POV150 < 0, NA, F_POV150),
         F_AGE17 = if_else(F_POV150 < 0, NA, F_POV150),
         F_DISABL = if_else(F_POV150 < 0, NA, F_POV150),
         F_SNGPNT = if_else(F_POV150 < 0, NA, F_POV150),
         F_LIMENG = if_else(F_POV150 < 0, NA, F_POV150),   
         F_THEME2 = if_else(F_POV150 < 0, NA, F_POV150),
         F_MINRTY = if_else(F_POV150 < 0, NA, F_POV150),
         F_THEME3 = if_else(F_POV150 < 0, NA, F_POV150),
         F_MUNIT = if_else(F_POV150 < 0, NA, F_POV150),
         F_MOBILE = if_else(F_POV150 < 0, NA, F_POV150),
         F_CROWD = if_else(F_POV150 < 0, NA, F_POV150),
         F_NOVEH = if_else(F_POV150 < 0, NA, F_POV150),
         F_GROUPQ = if_else(F_POV150 < 0, NA, F_POV150),
         F_THEME4 = if_else(F_POV150 < 0, NA, F_POV150),
         F_TOTAL = if_else(F_POV150 < 0, NA, F_POV150))

svi_county_EP_cols <- colnames(svi_county %>% select(contains("EP")))
svi_county[svi_county_EP_cols] <- svi_county[svi_county_EP_cols] / 100

colnames(svi_county)[4:74] <- paste0("SVI_", colnames(svi_county)[4:74])
```

#######################################################################
############### County Health Rankings & Roadmaps (CHR) ###############
#######################################################################
# save updated csv file and data dictionary. ensure column names match.

# Import and clean CHR data
```{r import and clean CHR data}
library(janitor)
chr_file <- read.csv("F:\\Data_Science\\SDoH\\sdoh 2025 update\\2025 CHR CSV Analytic Data.csv", header = FALSE)[-1,] %>%
  row_to_names(row_number = 1)

chr_dictionary <- data.frame(rbind(
  c("v036_rawvalue", "poor physical health days", "poor_physical_health_days"),
  c("v042_rawvalue", "poor mental health days", "poor_mental_health_days"),
  c("v132_rawvalue", "access to exercise opportunities", "prop_exercise_access"),
  c("v133_rawvalue", "food environment index", "food_environ_index"),
  c("v085_rawvalue", "uninsured", "prop_uninsured"),
  c("v136_rawvalue", "severe housing problems", "prop_severe_housing_probs"),
  c("v136_other_data_1", "percentage of households with high housing costs", "prop_high_housing_cost"),
  c("v136_other_data_2", "percentage of households with overcrowding", "prop_overcrowding"),
  c("v136_other_data_3", "percentage of households with lack of kitchen or plumbing facilities", "prop_no_kitchen_plumbing"),
  c("v125_rawvalue", "air pollution: particulate matter", "air_pollution"),
  c("v124_rawvalue", "drinking water violations", "drinking_water_violations"),
  c("v166_rawvalue", "broadband access", "prop_broadband_access"),
  c("v069_rawvalue", "some college", "prop_some_college"),
  c("v168_rawvalue", "high school completion", "prop_high_school_diploma"),
  c("v023_rawvalue", "unemployment", "prop_unemployment"),
  c("v044_rawvalue", "income inequality", "income_inequality"),
  c("v024_rawvalue", "children in poverty", "prop_children_in_poverty"),
  c("v140_rawvalue", "social associations", "social_associations"),
  c("v147_rawvalue", "life expectancy", "life_expectancy"),
  c("v127_rawvalue", "premature age-adjusted mortality", "premature_mortality"),
  c("v060_rawvalue", "diabetes prevalence", "diabetes_prevalence"),
  c("v011_rawvalue", "adult obesity", "prop_adult_obesity"),
  c("v083_rawvalue", "limited access to healthy foods", "prop_limited_healthy_food_access"),
  c("v139_rawvalue", "food insecurity", "prop_food_insecurity"),
  c("v009_rawvalue", "adult smoking", "prop_adult_smoking"),
  c("v070_rawvalue", "physical inactivity", "prop_physical_inactivity"),
  c("v154_rawvalue", "severe housing cost burden", "prop_severe_housing_cost"),
  c("v063_rawvalue", "median household income", "median_household_income"),
  c("v058_rawvalue", "proportion rural", "prop_rural")
))
colnames(chr_dictionary) <- c("variable", "description", "name")

chr <- chr_file %>%
  select(fipscode,
         all_of(chr_dictionary$variable)) %>%
  mutate(across(everything(), as.numeric),
         fipscode = as.character(fipscode))
colnames(chr) <- c("FIPS", chr_dictionary$name)

colnames(chr)[-1] <- paste0("CHR_", colnames(chr)[-1])
```

#######################################################
############### Census Total Population ###############
#######################################################

# import and clean total population
```{r import and clean total population}
total_pop_county_file <- read.csv("F:\\Data_Science\\SDoH\\Data\\sdoh 2024\\CB_Total Population Table P1 - County 2020.csv")

total_pop_county <- total_pop_county_file[-1,]
total_pop_county[,5:6] <- str_split_fixed(total_pop_county$GEO_ID, "S", 2) 
total_pop_county <- total_pop_county %>%
  mutate(Label = X,
         FIPS = as.character(as.numeric(V6)),
         TOT_POP = as.numeric(P1_001N)) %>%
  filter(Label=="0500000U") %>%
  select(FIPS,
         TOT_POP) 
```

#####################################################################
############### Create County-Level Aggregate Table #################
#####################################################################

```{r create county table}
table_county <- nchs %>%
  left_join(acs_county, by="FIPS") %>%
  left_join(cdc_county, by="FIPS") %>%
  left_join(svi_county, by="FIPS") %>%
  left_join(chr, by=c("FIPS")) %>%
  left_join(total_pop_county, by="FIPS") %>%
  relocate(STATE, .after=FIPS) %>%
  relocate(COUNTY, .after=STATE) %>%
  relocate(TOT_POP, .after=COUNTY) 

# save as RDS
# saveRDS(table_county, file = "F:\\Data_Science\\SDoH\\sdoh 2025 update\\SDoH_county_table 20250407.RDS")
```

```{r EDA for zip and county tables}
get_EDA <- function(data) {
  
  # structure
  print(str(data))
  
  # table with missing counts and frequency
  missing_summary <- data.frame(
    column = colnames(data),
    missing_count = sapply(data, function(x) sum(is.na(x))),
    missing_percentage = round(sapply(data, function(x) mean(is.na(x)) * 100), 2)
    ) %>%
    arrange(desc(missing_percentage))
 print(missing_summary) 
 
  # bar graph: % of missing values by feature
  ggplot(missing_summary, aes(x = reorder(column, -missing_percentage), y = missing_percentage)) +
        geom_bar(stat = "identity", fill = "darksalmon") +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        labs(title = "Percentage of Missing Values by Variable",
             x = "Variables",
             y = "Percentage Missing") +
        scale_y_continuous(labels = scales::number_format(accuracy = 0.01))
}
get_EDA(table_zip)
get_EDA(table_county)

# compare feature distributions
for (col in colnames(table_zip)[-(1:3)]) {
 
  histogram <- ggplot(table_zip, aes(x = .data[[col]])) +
    geom_histogram(position = "identity", binwidth = 0.02, fill = "lightblue", color = "black") +
    labs(x = col, y = "Count", title = paste("Histogram of", col)) +
    theme_minimal()

  print(histogram)
}

# ACS median HH income
ggplot(table_zip, aes(x = ACS_median_HH_income)) +
    geom_histogram(position = "identity", binwidth = 1000, fill = "lightblue", color = "black") +
    labs(x = "ACS_median_HH_income", y = "Count", title = paste("Histogram of", "ACS_median_HH_income")) +
    theme_minimal()

for (col in colnames(table_county)[-(1:3)]) {
 
  histogram <- ggplot(table_county, aes(x = .data[[col]])) +
    geom_histogram(position = "identity", binwidth = 0.02, fill = "lightblue", color = "black") +
    labs(x = col, y = "Count", title = paste("Histogram of", col)) +
    theme_minimal()

  print(histogram)
}

# ACS median HH income
ggplot(table_county, aes(x = ACS_median_HH_income)) +
    geom_histogram(position = "identity", binwidth = 1000, fill = "lightblue", color = "black") +
    labs(x = "ACS_median_HH_income", y = "Count", title = paste("Histogram of", "ACS_median_HH_income")) +
    theme_minimal()

# CHR median HH income
ggplot(table_county, aes(x = CHR_median_household_income)) +
    geom_histogram(position = "identity", binwidth = 1000, fill = "lightblue", color = "black") +
    labs(x = "CHR_median_household_income", y = "Count", title = paste("Histogram of", "CHR_median_household_income")) +
    theme_minimal()

# total population
ggplot(table_county, aes(x = TOT_POP)) +
    geom_histogram(position = "identity", binwidth = 100000, fill = "lightblue", color = "black") +
    labs(x = "TOT_POP", y = "Count", title = paste("Histogram of", "TOT_POP")) +
    theme_minimal()
```

############################################################
############### Create Time Series Dataset #################
############################################################

```{r import and clean 2024 CHR dat}
chr_file_2024 <- read.csv("F:\\Data_Science\\SDoH\\sdoh 2025 update\\2024 CHR CSV Analytic Data.csv", header = FALSE)[-1,] %>%
  row_to_names(row_number = 1)

chr_dictionary_2024 <- data.frame(rbind(
  c("v036_rawvalue", "poor physical health days", "poor_physical_health_days"),
  c("v042_rawvalue", "poor mental health days", "poor_mental_health_days"),
  c("v132_rawvalue", "access to exercise opportunities", "prop_exercise_access"),
  c("v133_rawvalue", "food environment index", "food_environ_index"),
  c("v085_rawvalue", "uninsured", "prop_uninsured"),
  c("v136_rawvalue", "severe housing problems", "prop_severe_housing_probs"),
  c("v136_other_data_1", "percentage of households with high housing costs", "prop_high_housing_cost"),
  c("v136_other_data_2", "percentage of households with overcrowding", "prop_overcrowding"),
  c("v136_other_data_3", "percentage of households with lack of kitchen or plumbing facilities", "prop_no_kitchen_plumbing"),
  c("v125_rawvalue", "air pollution: particulate matter", "air_pollution"),
  c("v124_rawvalue", "drinking water violations", "drinking_water_violations"),
  c("v166_rawvalue", "broadband access", "prop_broadband_access"),
  c("v069_rawvalue", "some college", "prop_some_college"),
  c("v168_rawvalue", "high school completion", "prop_high_school_diploma"),
  c("v023_rawvalue", "unemployment", "prop_unemployment"),
  c("v044_rawvalue", "income inequality", "income_inequality"),
  c("v024_rawvalue", "children in poverty", "prop_children_in_poverty"),
  c("v140_rawvalue", "social associations", "social_associations"),
  c("v147_rawvalue", "life expectancy", "life_expectancy"),
  c("v127_rawvalue", "premature age-adjusted mortality", "premature_mortality"),
  c("v060_rawvalue", "diabetes prevalence", "diabetes_prevalence"),
  c("v011_rawvalue", "adult obesity", "prop_adult_obesity"),
  c("v083_rawvalue", "limited access to healthy foods", "prop_limited_healthy_food_access"),
  c("v139_rawvalue", "food insecurity", "prop_food_insecurity"),
  c("v009_rawvalue", "adult smoking", "prop_adult_smoking"),
  c("v070_rawvalue", "physical inactivity", "prop_physical_inactivity"),
  c("v154_rawvalue", "severe housing cost burden", "prop_severe_housing_cost"),
  c("v063_rawvalue", "median household income", "median_household_income"),
  c("v058_rawvalue", "proportion rural", "prop_rural")
))
colnames(chr_dictionary_2024) <- c("variable", "description", "name")

chr_2024 <- chr_file_2024 %>%
  select(fipscode,
         all_of(chr_dictionary$variable)) %>%
  mutate(across(everything(), as.numeric),
         fipscode = as.character(fipscode))
colnames(chr_2024) <- c("FIPS", chr_dictionary_2024$name)

colnames(chr_2024)[-1] <- paste0("CHR_", colnames(chr_2024)[-1])
```

```{r create time series dataset}
# import 2024 tables
zip_2024 <- readRDS("F:\\Data_Science\\SDoH\\Tables\\master_zip_20240103.rds") 
county_2024 <- readRDS("F:\\Data_Science\\SDoH\\Tables\\master_county_20240702.rds")  %>%
  # add 2024 CHR data to 2024 dataset
  left_join(chr_2024, by = "FIPS")

# import 2025 tables
zip_2025 <- readRDS("F:\\Data_Science\\SDoH\\sdoh 2025 update\\SDoH_zip_table 20250402.RDS") 
county_2025 <- readRDS("F:\\Data_Science\\SDoH\\sdoh 2025 update\\SDoH_county_table 20250407.RDS") 

# select column names that are common across all datasets
zip_common_cols <- c("VERSION", intersect(names(zip_2024), names(zip_2025)))
county_common_cols <-  c("VERSION", intersect(names(county_2024), names(county_2025)))

# select column names that are new
zip_new_cols <- colnames(zip_2025)[!colnames(zip_2025) %in% colnames(zip_2024)]
county_new_cols <- colnames(county_2025)[!colnames(county_2025) %in% colnames(county_2024)]

# standardize zip and county tables to have common columns
zip_common_2024 <- zip_2024 %>% 
  mutate(VERSION = "2024") %>%
  select(all_of(zip_common_cols)) 
  
zip_common_2025 <- zip_2025 %>%
  mutate(VERSION = "2025") %>%
  select(all_of(zip_common_cols)) 

zip_common <- rbind(zip_common_2024, zip_common_2025)

county_common_2024 <- county_2024 %>% 
  mutate(VERSION = "2024") %>%
  select(all_of(county_common_cols))

county_common_2025 <- county_2025 %>%
  mutate(VERSION = "2025") %>%
  select(all_of(county_common_cols)) 

county_common <- rbind(county_common_2024, county_common_2025)

# writeClipboard(colnames(zip_common))
# writeClipboard(colnames(county_common))

# saveRDS(zip_common, "F:\\Data_Science\\SDoH\\sdoh 2025 update\\SDoH_zip_timeseries 20250407.RDS")
# saveRDS(county_common, "F:\\Data_Science\\SDoH\\sdoh 2025 update\\SDoH_county_timeseries 20250407.RDS")
```

```{r EDA for time series tables}
get_EDA(zip_common)
get_EDA(county_common)

# compare feature distributions
for (col in colnames(zip_common)[-(1:3)]) {
 
  histogram <- ggplot(zip_common, aes(x = .data[[col]], fill = factor(VERSION))) +
    geom_histogram(position = "identity", binwidth = 0.02, alpha = 0.7) +
    labs(x = col, y = "Count", fill = "Version", title = paste("Histogram of", col)) +
    theme_minimal()

  print(histogram)
}

# ACS median HH income
ggplot(zip_common, aes(x = ACS_median_HH_income, fill = factor(VERSION))) +
    geom_histogram(position = "identity", binwidth = 1000, alpha = 0.7) +
    labs(x = "ACS_median_HH_income", y = "Count", fill = "Version", title = paste("Histogram of", "ACS_median_HH_income")) +
    theme_minimal()

for (col in colnames(county_common)[-(1:4)]) {
 
  histogram <- ggplot(county_common, aes(x = .data[[col]], fill = factor(VERSION))) +
    geom_histogram(position = "identity", binwidth = 0.02, alpha = 0.7) +
    labs(x = col, y = "Count", fill = "Version", title = paste("Histogram of", col)) +
    theme_minimal()

  print(histogram)
}

# ACS median HH income
ggplot(county_common, aes(x = ACS_median_HH_income, fill = factor(VERSION))) +
    geom_histogram(position = "identity", binwidth = 1000, alpha = 0.7) +
    labs(x = "ACS_median_HH_income", y = "Count", fill = "Version", title = paste("Histogram of", "ACS_median_HH_income")) +
    theme_minimal()

# CHR median HH income
ggplot(county_common, aes(x = CHR_median_household_income, fill = factor(VERSION))) +
    geom_histogram(position = "identity", binwidth = 1000, alpha = 0.7) +
    labs(x = "CHR_median_household_income", y = "Count", fill = "Version", title = paste("Histogram of", "CHR_median_household_income")) +
    theme_minimal()

# total population
ggplot(county_common, aes(x = TOT_POP, fill = factor(VERSION))) +
    geom_histogram(position = "identity", binwidth = 100000, alpha = 0.7) +
    labs(x = "TOT_POP", y = "Count", fill = "Version", title = paste("Histogram of", "TOT_POP")) +
    theme_minimal()
```









```{r EDA for ACS}


# table with missing counts and frequency
missing_summary <- data.frame(
  column = colnames(zip_common_ACS_clean),
  missing_count = sapply(zip_common_ACS_clean, function(x) sum(is.na(x))),
  missing_percentage = round(sapply(zip_common_ACS_clean, function(x) mean(is.na(x)) * 100), 2)
  ) %>%
  arrange(desc(missing_percentage))
missing_summary

# bar graph: % of missing values by feature
ggplot(missing_summary, aes(x = reorder(column, -missing_percentage), y = missing_percentage)) +
      geom_bar(stat = "identity", fill = "darksalmon") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      labs(title = "Percentage of Missing Values by Variable",
           x = "Variables",
           y = "Percentage Missing") +
      scale_y_continuous(labels = scales::number_format(accuracy = 0.01))
```




